Regression models for Cylindrical data in Psychology

### 2019-03-08 Manuscript/Manuscript.tex
Manuscript update
Manuscript Update

\documentclass[man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Regression models for Cylindrical data in Psychology},
            pdfauthor={Jolien Cremers, Helena J.M. Pennings, \& Christophe Ley},
            pdfkeywords={cylindrical data, regression, interpersonal behavior},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\shorttitle{Cylindrical data in Psychology}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Department of Methodology and Statistics, Utrecht University\\\textsuperscript{2} TNO\\\textsuperscript{3} Department of Education, Utrecht University\\\textsuperscript{4} Department of Applied Mathematics, Computer Science and Statistics, Ghent University}
\keywords{cylindrical data, regression, interpersonal behavior}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

Correspondence concerning this article should be addressed to Jolien Cremers, . E-mail: \href{mailto:joliencremers@gmail.com}{\nolinkurl{joliencremers@gmail.com}}}

@@

Hernandez-Stumpfhauser, Breidt, and Woerd (2016) for the regression of the circular outcome. A detailed

@@

\leavevmode\hypertarget{ref-hernandez2016general}{}%
Hernandez-Stumpfhauser, D., Breidt, F. J., \& Woerd, M. J. van der. (2016). The general projected normal distribution of arbitrary dimension: Modeling and Bayesian inference. \emph{Bayesian Analysis}, \emph{12}(1), 113--133. doi:\href{https://doi.org/10.1214/15-BA989}{10.1214/15-BA989}

>>>>

\documentclass[man,mask]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Regression models for Cylindrical data in Psychology},
            pdfauthor={Jolien Cremers, Helena J.M. Pennings, \& Christophe Ley},
            pdfkeywords={cylindrical data, regression, interpersonal behavior},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\shorttitle{Cylindrical data in Psychology}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Department of Biostatistics, University of Copenhagen\\\textsuperscript{2} TNO\\\textsuperscript{3} Department of Education, Utrecht University\\\textsuperscript{4} Department of Applied Mathematics, Computer Science and Statistics, Ghent University}
\keywords{cylindrical data, regression, interpersonal behavior}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

Correspondence concerning this article should be addressed to Jolien Cremers, . E-mail: \href{mailto:jcre@sund.ku.dk}{\nolinkurl{jcre@sund.ku.dk}}}

@@

Hernandez-Stumpfhauser, Breidt, and \VANDER{Woerd}{Van der}{van der} Woerd (2016) for the regression of the circular outcome. A detailed

@@

\leavevmode\hypertarget{ref-hernandez2016general}{}%
Hernandez-Stumpfhauser, D., Breidt, F. J., \& \VANDER{Woerd}{Van der}{van der} Woerd, M. J. (2016). The general projected normal distribution of arbitrary dimension: Modeling and Bayesian inference. \emph{Bayesian Analysis}, \emph{12}(1), 113--133. doi:\href{https://doi.org/10.1214/15-BA989}{10.1214/15-BA989}

### 2019-03-08 Manuscript/SupplementaryMaterial.tex
Supplementary Material Update
Manuscript Update

\doublespacing

@@

Following Mastrantonio, Maruotti, \& Jona-Lasinio (2015) we consider in
this section two models where the relation between
\(\Theta \in [0, 2\pi)\) and \(Y\in (-\infty, + \infty)\) and \(q\)
covariates is specified as \begin{equation}\label{circlinlink}
Y = \gamma_0 + \gamma_{cos}*\cos(\Theta)*R + \gamma_{sin}*\sin(\Theta)*R + \gamma_1*x_1 + \dots + \gamma_q*x_q +  \epsilon,
\end{equation} \noindent where the random variable \(R\geq0\) will be
introduced below, the error term \(\epsilon \sim N(0, \sigma^2)\) with
variance \(\sigma^2>0\),
\(\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q\) are
the intercept and regression coefficients and \(x_1, \dots, x_q\) are
the \(q\) covariates. In both of these models the conditional
distribution of \(Y\) given \(\Theta=\theta\) and \(R = r\) is given by
\begin{equation}\label{ycondtheta}
f(y \mid \theta, r) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left[-\frac{(y - (\gamma_0 + \gamma_1x_1 + \dots + \gamma_qx_q+c))^{2}}{2\sigma^2}\right],\nonumber
\end{equation} \noindent where
\(c = \begin{bmatrix} r \cos(\theta) \\ r\sin(\theta) \end{bmatrix}^t \begin{bmatrix} \gamma_{cos} \\ \gamma_{sin} \end{bmatrix}\),
\(r \geq 0\). The linear outcome thus has a normal distribution
conditional on \(\Theta\) and \(R\) and contains already linear
covariates \(x_1, \dots, x_q\) in its location part.\newline \indent For
the circular outcome we assume either a projected normal (PN) or a
general projected normal (GPN) distribution. These distributions arise
from the radial projection of a distribution defined on the plane onto
the circle. The relation between a bivariate vector \(\boldsymbol{S}\)
in the plane and the circular outcome \(\Theta\) is defined as follows
\begin{equation}\label{projection}
\boldsymbol{S} = \begin{bmatrix} S^{I} \\ S^{II} \end{bmatrix} = R\boldsymbol{u} = \begin{bmatrix} R \cos (\Theta) \\  R\sin (\Theta) \end{bmatrix},
\end{equation} \noindent where \(R = \mid\mid \boldsymbol{S} \mid\mid\),
the Euclidean norm of the bivariate vector \(\boldsymbol{S}\). In the PN
distribution we assume
\(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{I})\) and in the
GPN we assume
\(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{\Sigma})\) where
\(\boldsymbol{\mu} \in \mathbb{R}^2\),
\(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \xi^2 & \xi\\ \xi & 1 \end{bmatrix}\),
and \(\xi,\tau \in (-\infty, +\infty)\) (as in Hernandez-Stumpfhauser,
Breidt, \& Woerd (2016)). This leads to the circular-linear PN (CL-PN)
and circular-linear GPN (CL-GPN) distributions. We will now detail how
we modify both cylindrical distributions to also incorporate covariates
for the circular part.

@@

$$l(\theta) =  \log(1) - \log(2\pi+\tau) - \sum \log(r_i) + (u_i^t\hat{\mu}_i\Sigma^{-1}(u_i^t\hat{\mu}_i)^t)/2\tau^2$$
where $u_i = (cos\theta_i, \sin \theta_i)$ and $\hat{\mu}_i = (\beta_0^{I} + \beta_0^{I}\text{SE}_i, \beta_0^{II} + \beta_0^{II}\text{SE}_i)^t$.

>>>>

\doublespacing
\DeclareRobustCommand{\VANDER}[3]{#2}
\DeclareRobustCommand{\VAN}[3]{#2}
\DeclareRobustCommand{\DEN}[3]{#2}

@@

Following Mastrantonio, Maruotti, \& Jona-Lasinio (2015) we consider in
this section two models where the relation between
\(\Theta \in [0, 2\pi)\) and \(Y\in (-\infty, + \infty)\) and \(q\)
covariates is specified as \begin{equation}\label{circlinlink}
Y = \gamma_0 + \gamma_{cos}*\cos(\Theta)*R + \gamma_{sin}*\sin(\Theta)*R + \gamma_1*x_1 + \dots + \gamma_q*x_q +  \epsilon,
\end{equation} \noindent where the random variable \(R\geq0\) will be
introduced below, the error term \(\epsilon \sim N(0, \sigma^2)\) with
variance \(\sigma^2>0\),
\(\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q\) are
the intercept and regression coefficients and \(x_1, \dots, x_q\) are
the \(q\) covariates. In both of these models the conditional
distribution of \(Y\) given \(\Theta=\theta\) and \(R = r\) is given by
\begin{equation}\label{ycondtheta}
f(y \mid \theta, r) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left[-\frac{(y - (\gamma_0 + \gamma_1x_1 + \dots + \gamma_qx_q+c))^{2}}{2\sigma^2}\right],\nonumber
\end{equation} \noindent where
\(c = \begin{bmatrix} r \cos(\theta) \\ r\sin(\theta) \end{bmatrix}^t \begin{bmatrix} \gamma_{cos} \\ \gamma_{sin} \end{bmatrix}\),
\(r \geq 0\). The linear outcome thus has a normal distribution
conditional on \(\Theta\) and \(R\) and contains already linear
covariates \(x_1, \dots, x_q\) in its location part.\newline \indent For
the circular outcome we assume either a projected normal (PN) or a
general projected normal (GPN) distribution. These distributions arise
from the radial projection of a distribution defined on the plane onto
the circle. The relation between a bivariate vector \(\boldsymbol{S}\)
in the plane and the circular outcome \(\Theta\) is defined as follows
\begin{equation}\label{projection}
\boldsymbol{S} = \begin{bmatrix} S^{I} \\ S^{II} \end{bmatrix} = R\boldsymbol{u} = \begin{bmatrix} R \cos (\Theta) \\  R\sin (\Theta) \end{bmatrix},
\end{equation} \noindent where \(R = \mid\mid \boldsymbol{S} \mid\mid\),
the Euclidean norm of the bivariate vector \(\boldsymbol{S}\). In the PN
distribution we assume
\(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{I})\) and in the
GPN we assume
\(\boldsymbol{S} \sim N_2(\boldsymbol{\mu}, \boldsymbol{\Sigma})\) where
\(\boldsymbol{\mu} \in \mathbb{R}^2\),
\(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \xi^2 & \xi\\ \xi & 1 \end{bmatrix}\),
and \(\xi,\tau \in (-\infty, +\infty)\) (as in Hernandez-Stumpfhauser,
Breidt, \& \VANDER{Woerd}{Van der}{van der} Woerd (2016)). This leads to
the circular-linear PN (CL-PN) and circular-linear GPN (CL-GPN)
distributions. We will now detail how we modify both cylindrical
distributions to also incorporate covariates for the circular part.

@@

$$l(\theta, r) =  \log(1) - \log(2\pi+\tau) - \sum \log(r_i) + (u_i^t\hat{\mu}_i\Sigma^{-1}(u_i^t\hat{\mu}_i)^t)/2\tau^2$$
where $u_i = (cos\theta_i, \sin \theta_i)$ and $\hat{\mu}_i = (\beta_0^{I} + \beta_0^{I}\text{SE}_i, \beta_0^{II} + \beta_0^{II}\text{SE}_i)^t$.
